Finding a closest point in a lattice of Voronoi's first kind

### 2014-01-14 paper.tex
minor
added more edits.  Looking good

The paper is structured as follows.  Section~\ref{sec:voron-cells-relev} describes the relevant vectors and the \emph{Voronoi cell} of a lattice. Section~\ref{sec:iterative-slicer} describes a procedure to find a closest lattice point by enumerating a series of relevant vectors.  The series is guaranteed to converge to a closest lattice point after a finite number of terms.  In general the procedure might be computationally expensive because the number of terms required might be large and because computation of each relevant vector in the series might be expensive.  Section~\ref{sec:latt-voron-first} describes lattices of Voronoi's first kind and their obtuse superbasis.  It is shown that for these lattices, the series of relevant vectors results in a closest lattice point after at most $n$ terms.  Section~\ref{sec:comp-clos-relev} shows that each relevant vector in the series can be computed in $O(n^3)$ operations by computing a minimum cut in an undirected flow network.  Section~\ref{sec:discussion} discusses some potential applications of this algorithm and poses some interesting questions for future research.

@@

\begin{proposition}\label{eq:latticepointsinvorcvell}
Let $\Lambda \subset \reals^{m}$ be an $n$-dimensional lattice.  For $r\in \reals$ let $\ceil{r}$ denote the smallest integer strictly larger than $r$.  Let $t \in \reals^m$.  The number of lattice points inside the scaled and translated Voronoi cell $r \vor(\Lambda) + t$ is atmost $\ceil{r}^n$.
\end{proposition}
\begin{proof}
It is convenient to modify the boundary of the Voronoi cell so that it tesselates $\reals^m$ under translations by $\Lambda$.  With this aim we let $V \subset \vor(\Lambda)$ contain all those points from the interior of $\vor(\Lambda)$, but have opposing faces open and closed.  That is, if $x \in V$ is on the boundary of $V$ then $-x \notin V$.  With this definition it can be asserted that $V$ tesslates $\reals^m$ under translations by $\Lambda$.  That is, $\reals^{m} = \cup_{x \in \Lambda}(V + x)$ and the intersection $(V + x)\cap(V+y)$ is empty for distinct lattice points $x$ and $y$~\cite[Sec.~2.2]{McKilliam2010thesis}.  Now, for positive integer $k$, the scaled and translated cell $kV + t$ contain precisely one of each of the coset representatives of the group quotient $\Lambda/k\Lambda$~\cite[Sec.~2.4]{McKilliam2010thesis}.  There are $k^n$ coset representatives.  Thus, the number of lattice point inside $r \vor(\Lambda) + t \subset \ceil{r}V + t$ is at most $\ceil{r}^n$.
\end{proof}

\begin{proposition}\label{eq:latticepointsinsphere}
Let $\Lambda \subset \reals^{m}$ be an $n$-dimensional lattice with packing radius $\rho$.  Let $S$ be an $m$-dimensional hypersphere of radius $r$ centered at $t \in \reals^{m}$.  The number of lattice points from $\Lambda$ in the sphere $S$ is less than $\ceil{r/\rho}^{n}$ where $\ceil{\cdot}$ denotes the smallest integer strictly larger than its argument. 
 \end{proposition}
 \begin{proof}
The packing radius $\rho$ is the Euclidean length of a point on the boundary of the Voronoi cell $\vor(\Lambda)$ that is closest to the origin. Therefore, the sphere $S$ is a subset of $\vor(\Lambda)$ scaled by $r/\rho$ and translated by $t$.  That is, $S \subset r/\rho \vor(\Lambda) + t$.  The proof follows because the number of lattice points in $r/\rho \vor(\Lambda) + t$ is less than $\ceil{r/\rho}^n$ by Proposition~\ref{eq:latticepointsinvorcvell}.
\end{proof}

@@

Classical examples of lattices of Voronoi's first kind are the $n$ dimensional root lattice $A_n$ and its dual lattice $A_n^*$~\cite{SPLAG}.  These lattices have convenient bases represented in $n+1$ dimensional space.  An obtuse superbasis for $A_n$ is of the form
\begin{align*}
b_i &= e_i - e_{i+1}, \qquad i = 1, \dots, n \\
b_{n+1} &= e_{n+1} - e_{1},
\end{align*}
where $e_i$ is a vector of length $n+1$ with elements equal to zero except for the $i$th element that is equal to one.  The dual lattice $A_{n}^*$ is also of Voronoi's first kind.  The $i$th vector $b_i$ of the obtuse superbasis has $j$th element equal to
\[
b_{ij} = \begin{cases}
\frac{n}{n+1} & j = i \\
-\frac{1}{n+1} & j \neq i.
\end{cases}
\]
For $A_n$ and $A_n^*$ there exists efficient algorithms that can compute a closest lattice point in $O(n)$ operations~\cite{McKilliam2009CoxeterLattices,Conway1982FastQuantDec} so we do not recommend using the algorithm described in this paper for $A_n$ and $A_n^*$.  The fast algorithms for $A_n$ and $A_n^*$ exploit the special structure of these lattices and are not applicable to other lattices.  In contrast, the algorithm we describe here works for all lattices of Voronoi's first kind.

@@

\section{A series of relevant vectors from a lattice of Voronoi's first kind}

@@

\begin{lemma}\label{lem:decSellings}
Let $p \in \reals^{n+1}$ and let $S$ and $T$ be subsets of the indices of $p$.  The following equalities hold:
\begin{enumerate}
\item  $\|Bp\|^2 - \|B(p + \onebf_S)\|^2 = \Phi(S, p)$, \label{eq:lem:decSellingsinc}
\item  $\|Bp\|^2 - \|B(p - \onebf_S)\|^2 = \Phi(\bar{S}, p)$, \label{eq:lem:decSellingsdec}
\item  ${\displaystyle \|Bp\|^2 - \|B(p + \onebf_S - \onebf_T)\|^2 = \Phi(S, p) + \Phi(\bar{T},p) + \sum_{i\in S}\sum_{j\in T} q_{ij} }$.  \label{eq:lem:decSellingsincdec}
\end{enumerate}
\end{lemma}
\begin{proof}
Part~\ref{eq:lem:decSellingsincdec} follows immediately from parts~\ref{eq:lem:decSellingsinc} and~\ref{eq:lem:decSellingsdec} because
\begin{align*}
\|Bp\|^2 - &\|B(p + \onebf_S - \onebf_T)\|^2 \\
&= \|Bp\|^2-\|B(p + \onebf_S)\|^2 +  \|Bp\|^2-\|B(p - \onebf_T)\|^2 + \sum_{i\in S}\sum_{j\in T} q_{ij}.
\end{align*}
We give a proof for part~\ref{eq:lem:decSellingsinc}.  The proof for part~\ref{eq:lem:decSellingsdec} is similar.  
Put $Q = B^\prime B$ where supercript $^\prime$ indicates the vector or matrix transpose.  The $n+1$ by $n+1$ matrix $Q$ has elements given by the Selling parameters, that is, $Q_{ij} = q_{ij}=\dotprod{b_i}{b_j}$.  Denote by $\onebf$ the column vector of length $n+1$ containing all ones.  Now $B\onebf = \sum_{i=1}^{n+1}b_i = \zerobf$ as a result of the superbasis condition~\eqref{eq:superbasecond} and so $Q\onebf = \zerobf$.  Since $\onebf_S = \onebf - \onebf_{\bar{S}}$ it follows that $Q\onebf_S = -Q\onebf_{\bar{S}}$.  With $\circ$ the elementwise vector product, i.e., the Schur or Hadamard product, we have
 \begin{align*}
 \|Bp\|^2 - \|B(p + \onebf_S)\|^2 &= -\onebf^\prime_S Q \onebf_S - 2 p^\prime Q \onebf_S \\
 &= \onebf^\prime_S Q \onebf_{\bar{S}} - 2 p^\prime Q \onebf_S \\
&= \onebf^\prime_S Q \onebf_{\bar{S}} - 2 (p\circ\onebf_{\bar{S}})^\prime Q \onebf_S - 2 (p\circ\onebf_S)^\prime Q \onebf_S \\
&= \onebf^\prime_S Q \onebf_{\bar{S}} - 2 (p\circ\onebf_{\bar{S}})^\prime Q \onebf_{S} + 2 (p\circ\onebf_S)^\prime Q \onebf_{\bar{S}}
 \end{align*}
which is precisely $\Phi(S,p)$.
% Observe that
% \[
% \sum_{i=1}^{n+1}\sum_{j=1}^{n+1} q_{ij} p_i^2 =  \sum_{i=1}^{n+1}p_i^2\sum_{j=1}^{n+1} q_{ij} = 0
% \]
% and
% \[
% \sum_{i=1}^{n+1}\sum_{j=1}^{n+1} q_{ij} p_j^2 =  \sum_{j=1}^{n+1}p_j^2\sum_{i=1}^{n+1} q_{ij} = 0
% \]
% since $\sum_{i=1}^{n+1} q_{ij} = \sum_{j=1}^{n+1} q_{ij} = 0$ as a result of the superbasis condition~\eqref{eq:superbasecond}.  Now
% \begin{align*}
% \|Bp\|^2 &= \sum_{i=1}^{n+1} \sum_{j=1}^{n+1} q_{ij} p_i p_j \\
% &= \sum_{i=1}^{n+1} \sum_{j=1}^{n+1} q_{ij} (\tfrac{1}{2}p_j^2 + p_i p_j -  \tfrac{1}{2}p_i^2) \\
% &= -\frac{1}{2} \sum_{i=1}^{n+1} \sum_{j=1}^{n+1} q_{ij} (p_i - p_j)^2.
% \end{align*}
% This is closely related to what is known as Sellings formula~\cite[Proposition 2.3.1]{Valentin2003_coverings_tilings_low_dimension}~\cite{Selling1874}.  Put $g = p - \onebf_S$.  It follows that
% \[
% \|Bp\|^2 - \|Bg\|^2 = \frac{1}{2} \sum_{i=1}^{n+1}\sum_{j=1}^{n+1}q_{ij}d_{ij},
% \]
% where $d_{ij} = (g_i - g_j)^2 - (p_i - p_j)^2$.  If both $i \notin S$ and $j \notin S$ then $p_i = g_i$, $p_j = g_j$ and so $d_{ij} = 0$.  Similarly if both $i \in S$ and $j \in S$ then 
% \[
% d_{ij} = (p_i-1 - p_j+1)^2 - (p_i - p_j)^2 = 0.
% \]
% If $i \in S$ and $j \notin S$ then
% \[
% d_{ij} = (p_i-1 - p_j)^2 - (p_i - p_j)^2= 1 + 2p_j - 2p_i,
% \]
% while if $i \notin S$ and $j \in S$ then
% \[
% d_{ij} = (p_i - p_j+1)^2 - (p_i - p_j)^2 = 1 + 2p_i - 2p_j.
% \]
% We thus have
% \begin{align*}
% \|Bp\|^2 - \|Bg\|^2 &= \frac{1}{2} \sum_{i \in S}\sum_{j \notin S}q_{ij}(1 + 2p_j - 2p_i) + \frac{1}{2} \sum_{i \notin S}\sum_{j \in S}q_{ij}(1 + 2p_i - 2p_j) \\
% &= \sum_{i \in S}\sum_{j \notin S}q_{ij}(1 + 2p_j - 2p_i),
% \end{align*}
% where the last line holds because $q_{ij} = q_{ji}$.
\end{proof}

@@

Put $r = z - u - h$.  By~\eqref{eq:hijvijieq} we have $r_i - r_j \geq p_{i} - p_{j}$ when either $i \in S$ or $j \in T$.  Now, since $q_{ij} \leq 0$ for $i \neq j$,
\[
\Phi(S,r) = \sum_{i \in S}\sum_{j \notin S}q_{ij}(1 + 2r_i - 2r_j) \leq \sum_{i \in S}\sum_{j \notin S}q_{ij}(1 + 2p_i - 2p_j) = \Phi(S,p),
\]
and
\[
\Phi(\bar{T},r) = \sum_{i \notin T}\sum_{j \in T}q_{ij}(1 + 2r_i - 2r_j) \leq \sum_{i \notin T}\sum_{j \in T}q_{ij}(1 + 2p_i - 2p_j) = \Phi(\bar{T},p).
\]
Using part~\ref{eq:lem:decSellingsincdec} of Lemma~\ref{lem:decSellings} again,
\begin{align*}
\|B(z-u-h)\|^2 - \|B(z-u-g)\|^2 &= \|Br\|^2 - \|B(r+\onebf_S - \onebf_T)\|^2 \\
&= \Phi(S,r) + \Phi(\bar{T},r) + \sum_{i\in S}\sum_{j \in S} q_{ij} \\
&\leq \Phi(S,p) + \Phi(\bar{T}, p) + \sum_{i\in S}\sum_{j \in S} q_{ij} < 0 
\end{align*}
as a result of~\eqref{eq:BpBpdecincineq}.  However, $h \in \{0,1\}^{n+1}$ and so this implies
\[
\|B(z-u-g)\|^2 > \|B(z-u-h)\|^2 \geq \min_{t \in \{0,1\}^{n+1}}\|B(z - u - t)\|^2
\]
contradicting~\eqref{eq:qismin}.  Thus, or original supposition is false and $B(u+g+w)$ is a closest lattice point to $y$. Because $\rng(w) = \ell-1$ the lattice point $B(u+g)$ is $(\ell-1)$-close to $y$.
 \end{proof}

@@

@@

A question of immediate interest is: do there exist lattices of Voronoi's first kind that produce good codes or good quantisers?  Since lattices that produce good codes and quantisers often also describe dense~\emph{sphere packings}~\cite{SPLAG}, a related question is: do there exist lattices of Voronoi's first kind that produce dense sphere packings?  These question do not appear to have trivial answers.  The questions have heightened importance due to the algorithm described in this paper.

>>>>

The paper is structured as follows.  Section~\ref{sec:voron-cells-relev} describes the relevant vectors and the \emph{Voronoi cell} of a lattice. Section~\ref{sec:iterative-slicer} describes a procedure to find a closest lattice point by enumerating a series of relevant vectors.  The series is guaranteed to converge to a closest lattice point after a finite number of terms.  In general the procedure might be computationally expensive because the number of terms required might be large and because computation of each relevant vector in the series might be expensive.  Section~\ref{sec:latt-voron-first} describes lattices of Voronoi's first kind and their obtuse superbasis.  In Section~\ref{sec:seri-relev-vect} it is shown that for these lattices, the series of relevant vectors results in a closest lattice point after at most $n$ terms.  Section~\ref{sec:comp-clos-relev} shows that each relevant vector in the series can be computed in $O(n^3)$ operations by computing a minimum cut in an undirected flow network.  Section~\ref{sec:discussion} discusses some potential applications of this algorithm and poses some interesting questions for future research.

@@

\begin{proposition}\label{eq:latticepointsinvorcvell}
Let $\Lambda \subset \reals^{m}$ be an $n$-dimensional lattice.  For $r\in \reals$ let $\ceil{r}$ denote the smallest integer strictly larger than $r$.  Let $t \in \reals^m$.  The number of lattice points inside the scaled and translated Voronoi cell $r \vor(\Lambda) + t$ is at most $\ceil{r}^n$.
\end{proposition}
\begin{proof}
It is convenient to modify the boundary of the Voronoi cell so that it tessellates $\reals^m$ under translations by $\Lambda$.  With this aim we let $V \subset \vor(\Lambda)$ contain all those points from the interior of $\vor(\Lambda)$, but have opposing faces open and closed.  That is, if $x \in V$ is on the boundary of $V$ then $-x \notin V$.  With this definition it can be asserted that $V$ tessellates $\reals^m$ under translations by $\Lambda$.  That is, $\reals^{m} = \cup_{x \in \Lambda}(V + x)$ and the intersection $(V + x)\cap(V+y)$ is empty for distinct lattice points $x$ and $y$~\cite[Sec.~2.2]{McKilliam2010thesis}.  Now, for positive integer $k$, the scaled and translated cell $kV + t$ contains precisely one coset representative for each element of the quotient group $\Lambda/k\Lambda$~\cite[Sec.~2.4]{McKilliam2010thesis}.  There are $k^n$ coset representatives.  Thus, the number of lattice points inside $r \vor(\Lambda) + t \subset \ceil{r}V + t$ is at most $\ceil{r}^n$.
\end{proof}

\begin{proposition}\label{eq:latticepointsinsphere}
Let $\Lambda \subset \reals^{m}$ be an $n$-dimensional lattice with packing radius $\rho$.  Let $S$ be an $m$-dimensional hypersphere of radius $r$ centered at $t \in \reals^{m}$.  The number of lattice points from $\Lambda$ in the sphere $S$ is at most $\ceil{r/\rho}^{n}$ where $\ceil{\cdot}$ denotes the smallest integer strictly larger than its argument. 
 \end{proposition}
 \begin{proof}
The packing radius $\rho$ is the Euclidean length of a point on the boundary of the Voronoi cell $\vor(\Lambda)$ that is closest to the origin. Therefore, the sphere $S$ is a subset of $\vor(\Lambda)$ scaled by $r/\rho$ and translated by $t$.  That is, $S \subset r/\rho \vor(\Lambda) + t$.  The proof follows because the number of lattice points in $r/\rho \vor(\Lambda) + t$ is at most $\ceil{r/\rho}^n$ by Proposition~\ref{eq:latticepointsinvorcvell}.
\end{proof}

@@

Classical examples of lattices of Voronoi's first kind are the $n$ dimensional root lattice $A_n$ and its dual lattice $A_n^*$~\cite{SPLAG}.  % These lattices have convenient bases represented in $n+1$ dimensional space.  An obtuse superbasis for $A_n$ is of the form
% \begin{align*}
% b_i &= e_i - e_{i+1}, \qquad i = 1, \dots, n \\
% b_{n+1} &= e_{n+1} - e_{1},
% \end{align*}
% where $e_i$ is a vector of length $n+1$ with elements equal to zero except for the $i$th element that is equal to one.  The dual lattice $A_{n}^*$ is also of Voronoi's first kind.  The $i$th vector $b_i$ of the obtuse superbasis has $j$th element equal to
% \[
% b_{ij} = \begin{cases}
% \frac{n}{n+1} & j = i \\
% -\frac{1}{n+1} & j \neq i.
% \end{cases}
% \]
For $A_n$ and $A_n^*$ there exist efficient algorithms that can compute a closest lattice point in $O(n)$ operations~\cite{McKilliam2009CoxeterLattices,Conway1982FastQuantDec}. For this reason we do not recommend using the algorithm described in this paper for $A_n$ and $A_n^*$.  
The fast algorithms for $A_n$ and $A_n^*$ rely of the special structure of these lattices and are not applicable to other lattices.  In contrast, the algorithm we describe here works for all lattices of Voronoi's first kind.  Questions that arise are: how ``large'' (in some sense) is the set of lattices of Voronoi's first kind?  Are there lattices of Voronoi's first kind that are useful in applications such as coding, quantisation, or signal processing?  We discuss these questions in Section~\ref{sec:discussion}.  We now focus on the problem of computing a closest lattice point in a lattice of Voronoi's first kind.

@@

\section{A series of relevant vectors from a lattice of Voronoi's first kind}\label{sec:seri-relev-vect}

@@

\begin{lemma}\label{lem:decSellings}
Let $p \in \reals^{n+1}$ and let $S$ and $T$ be subsets of the indices of $p$.  The following equalities hold:
\begin{enumerate}
\item  $\|Bp\|^2 - \|B(p + \onebf_S)\|^2 = \Phi(S, p)$, \label{eq:lem:decSellingsinc}
\item  $\|Bp\|^2 - \|B(p - \onebf_S)\|^2 = \Phi(\bar{S}, p)$, \label{eq:lem:decSellingsdec}
\item  ${\displaystyle \|Bp\|^2 - \|B(p + \onebf_S - \onebf_T)\|^2 = \Phi(S, p) + \Phi(\bar{T},p) + \sum_{i\in S}\sum_{j\in T} q_{ij} }$.  \label{eq:lem:decSellingsincdec}
\end{enumerate}
\end{lemma}
\begin{proof}
Part~\ref{eq:lem:decSellingsincdec} follows immediately from parts~\ref{eq:lem:decSellingsinc} and~\ref{eq:lem:decSellingsdec} because
\begin{align*}
\|Bp\|^2 - &\|B(p + \onebf_S - \onebf_T)\|^2 \\
&= \|Bp\|^2-\|B(p + \onebf_S)\|^2 +  \|Bp\|^2-\|B(p - \onebf_T)\|^2 + \sum_{i\in S}\sum_{j\in T} q_{ij}.
\end{align*}
We give a proof for part~\ref{eq:lem:decSellingsinc}.  The proof for part~\ref{eq:lem:decSellingsdec} is similar.  
Put $Q = B^\prime B$ where superscript $^\prime$ indicates the vector or matrix transpose.  The $n+1$ by $n+1$ matrix $Q$ has elements given by the Selling parameters, that is, $Q_{ij} = q_{ij}=\dotprod{b_i}{b_j}$.  Denote by $\onebf$ the column vector of length $n+1$ containing all ones.  Now $B\onebf = \sum_{i=1}^{n+1}b_i = \zerobf$ as a result of the superbasis condition~\eqref{eq:superbasecond} and so $Q\onebf = \zerobf$.  Since $\onebf_S = \onebf - \onebf_{\bar{S}}$ it follows that $Q\onebf_S = -Q\onebf_{\bar{S}}$.  With $\circ$ the elementwise vector product, i.e., the Schur or Hadamard product, we have
 \begin{align*}
 \|Bp\|^2 - \|B(p + \onebf_S)\|^2 &= -\onebf^\prime_S Q \onebf_S - 2 p^\prime Q \onebf_S \\
 &= \onebf^\prime_S Q \onebf_{\bar{S}} - 2 p^\prime Q \onebf_S \\
&= \onebf^\prime_S Q \onebf_{\bar{S}} - 2 (p\circ\onebf_{\bar{S}})^\prime Q \onebf_S - 2 (p\circ\onebf_S)^\prime Q \onebf_S \\
&= \onebf^\prime_S Q \onebf_{\bar{S}} - 2 (p\circ\onebf_{\bar{S}})^\prime Q \onebf_{S} + 2 (p\circ\onebf_S)^\prime Q \onebf_{\bar{S}}
 \end{align*}
which is precisely $\Phi(S,p)$.
% Observe that
% \[
% \sum_{i=1}^{n+1}\sum_{j=1}^{n+1} q_{ij} p_i^2 =  \sum_{i=1}^{n+1}p_i^2\sum_{j=1}^{n+1} q_{ij} = 0
% \]
% and
% \[
% \sum_{i=1}^{n+1}\sum_{j=1}^{n+1} q_{ij} p_j^2 =  \sum_{j=1}^{n+1}p_j^2\sum_{i=1}^{n+1} q_{ij} = 0
% \]
% since $\sum_{i=1}^{n+1} q_{ij} = \sum_{j=1}^{n+1} q_{ij} = 0$ as a result of the superbasis condition~\eqref{eq:superbasecond}.  Now
% \begin{align*}
% \|Bp\|^2 &= \sum_{i=1}^{n+1} \sum_{j=1}^{n+1} q_{ij} p_i p_j \\
% &= \sum_{i=1}^{n+1} \sum_{j=1}^{n+1} q_{ij} (\tfrac{1}{2}p_j^2 + p_i p_j -  \tfrac{1}{2}p_i^2) \\
% &= -\frac{1}{2} \sum_{i=1}^{n+1} \sum_{j=1}^{n+1} q_{ij} (p_i - p_j)^2.
% \end{align*}
% This is closely related to what is known as Sellings formula~\cite[Proposition 2.3.1]{Valentin2003_coverings_tilings_low_dimension}~\cite{Selling1874}.  Put $g = p - \onebf_S$.  It follows that
% \[
% \|Bp\|^2 - \|Bg\|^2 = \frac{1}{2} \sum_{i=1}^{n+1}\sum_{j=1}^{n+1}q_{ij}d_{ij},
% \]
% where $d_{ij} = (g_i - g_j)^2 - (p_i - p_j)^2$.  If both $i \notin S$ and $j \notin S$ then $p_i = g_i$, $p_j = g_j$ and so $d_{ij} = 0$.  Similarly if both $i \in S$ and $j \in S$ then 
% \[
% d_{ij} = (p_i-1 - p_j+1)^2 - (p_i - p_j)^2 = 0.
% \]
% If $i \in S$ and $j \notin S$ then
% \[
% d_{ij} = (p_i-1 - p_j)^2 - (p_i - p_j)^2= 1 + 2p_j - 2p_i,
% \]
% while if $i \notin S$ and $j \in S$ then
% \[
% d_{ij} = (p_i - p_j+1)^2 - (p_i - p_j)^2 = 1 + 2p_i - 2p_j.
% \]
% We thus have
% \begin{align*}
% \|Bp\|^2 - \|Bg\|^2 &= \frac{1}{2} \sum_{i \in S}\sum_{j \notin S}q_{ij}(1 + 2p_j - 2p_i) + \frac{1}{2} \sum_{i \notin S}\sum_{j \in S}q_{ij}(1 + 2p_i - 2p_j) \\
% &= \sum_{i \in S}\sum_{j \notin S}q_{ij}(1 + 2p_j - 2p_i),
% \end{align*}
% where the last line holds because $q_{ij} = q_{ji}$.
\end{proof}

@@

Put $r = z - u - h$.  By~\eqref{eq:hijvijieq} we have $r_i - r_j \geq p_{i} - p_{j}$ when either $i \in S$ or $j \in T$.  Now, since $q_{ij} \leq 0$ for $i \neq j$,
\[
\Phi(S,r) = \sum_{i \in S}\sum_{j \notin S}q_{ij}(1 + 2r_i - 2r_j) \leq \sum_{i \in S}\sum_{j \notin S}q_{ij}(1 + 2p_i - 2p_j) = \Phi(S,p),
\]
and
\[
\Phi(\bar{T},r) = \sum_{i \notin T}\sum_{j \in T}q_{ij}(1 + 2r_i - 2r_j) \leq \sum_{i \notin T}\sum_{j \in T}q_{ij}(1 + 2p_i - 2p_j) = \Phi(\bar{T},p).
\]
Using part~\ref{eq:lem:decSellingsincdec} of Lemma~\ref{lem:decSellings} again,
\begin{align*}
\|B(z-u-h)\|^2 - \|B(z-u-g)\|^2 &= \|Br\|^2 - \|B(r+\onebf_S - \onebf_T)\|^2 \\
&= \Phi(S,r) + \Phi(\bar{T},r) + \sum_{i\in S}\sum_{j \in S} q_{ij} \\
&\leq \Phi(S,p) + \Phi(\bar{T}, p) + \sum_{i\in S}\sum_{j \in S} q_{ij} < 0 
\end{align*}
as a result of~\eqref{eq:BpBpdecincineq}.  However, $h \in \{0,1\}^{n+1}$ and so this implies
\[
\|B(z-u-g)\|^2 > \|B(z-u-h)\|^2 \geq \min_{t \in \{0,1\}^{n+1}}\|B(z - u - t)\|^2
\]
contradicting~\eqref{eq:qismin}.  Thus, our original supposition is false and $B(u+g+w)$ is a closest lattice point to $y$. Because $\rng(w) = \ell-1$ the lattice point $B(u+g)$ is $(\ell-1)$-close to $y$.
 \end{proof}

@@

A question of immediate interest to communications engineers is: do there exist lattices of Voronoi's first kind that produce good codes or good quantisers?  Since lattices that produce good codes and quantisers often also describe dense~\emph{sphere packings}~\cite{SPLAG}, a related question is: do there exist lattices of Voronoi's first kind that produce dense sphere packings?  These question do not appear to have trivial answers.  The questions have heightened importance due to the algorithm described in this paper.

@@

A number applications such as phase unwrapping~\cite{Teunissen_GPS_1995,Hassibi_GPS_1998}, single frequency estimation~\cite{McKilliamFrequencyEstimationByPhaseUnwrapping2009}, and related signal processing problems~\cite{McKilliam2007,Clarkson2007,McKilliam2009IndentifiabliltyAliasingPolyphase,Quinn_sparse_noisy_SSP_2012} also require computing a closest lattice point.  In these applications the particular lattice arises from the signal processing problem under consideration.  If that lattice happens to be of Voronoi's first kind then our algorithm can be used.  An example where this occurs is the problem of computing the \emph{sample intrinsic mean} in circular statistics~\cite{McKilliam_mean_dir_est_sq_arc_length2010}.  In this particular problem the lattice $A_n^*$ is involved.  A fast closest point algorithm requiring only $O(n)$ operations exists for $A_n^*$~\cite{McKilliam2009CoxeterLattices,McKilliam2008b} and so the algorithm described in this paper is not needed in this particular case.  However, there many exist other signal processing problems where lattices of Voronoi's first kind arise.
